---
title: "permuteDE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{permuteDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

# Installation

------

permuteDE is designed to be run on Unix-based operating systems such as macOS and linux.

permuteDE installation currently requires `remotes` and `BiocManager` for installation of GitHub and Bioconductor packages. Run the following commands to install the various dependencies used by permuteDE:

First, install remotes (for installing GitHub packages) if it isn’t already installed:
```{r, eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
```

Then, install BiocManager (for installing bioconductor packages) if it isn’t already installed:
```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
```

Then, install permuteDE:
```{r, eval = FALSE}
remotes::install_github("corceslab/permuteDE", ref="main", repos = BiocManager::repositories(), upgrade = "never")
```

------

# Introduction

------

This tutorial provides a basic example of how to run permuteDE, an R package intended to help users assess which differential expression comparisons can be trusted and which should be met with skepticism. 

permuteDE is applicable to any sequencing data suitable for differential expression analysis. It takes as input a Seurat object, SingleCellExperiment object, or matrix (see [Advanced Options](<LINK>) for details about input object types). Detailed parameter definitions are available under the [Functions](<LINK>) tab.

Differential expression analyses are susceptible to false positives, and it can be difficult to prioritize the most robust results for further study. permuteDE is based on the premise that the differential expression results that are most likely to be validated in subsequent experiments come from comparisons that have a higher number of significant differentially expressed features than would be expected by chance. The use of permutation testing allows permuteDE to generate a distribution estimating how many false positive significant differentially expressed features occur by chance alone.

permuteDE proceeds in two main steps. First, the user will use function `runDE()` to run differential expression analysis using one of the several available methods. Second, these results are then passed to the `permuteDE()` function, which performs the permutation test. 

You’ll need the following packages installed to run this tutorial:
```{r, message = FALSE, warning = FALSE}
library(Seurat)
library(permuteDE)
```

### Input data

permuteDE takes as input a Seurat, SingleCellExperiment, or matrix. If a matrix is provided, it may be either a cell-level matrix or a pre-generated pseudobulk matrix. This tutorial uses a Seurat object; for details related to other object types, see [Advanced Options](<LINK>). permuteDE expects raw counts, and normalization is performed during the differential expression analysis depending on the selected method.

To demonstrate how to run permuteDE, we’ll use a small sample dataset consisting of mouse brain snRNA-seq data, stored as a Seurat object.

```{r, message = FALSE, warning = TRUE, fig.height = 5}
# Load dataset from permuteDE package
data("sample_data")

# Seurat object containing a count matrix, UMAP coordinates, and cell metadata
plotDimReduction(reduction = sample_data@reductions$umap@cell.embeddings,
                 split_labels = sample_data$cell_type,
                 color_by = "split")
```

------

# Running permuteDE

------

permuteDE proceeds in two main steps, each corresponding to a function: (1) `runDE()` and (2) `permuteDE()`.

## Step 1: `runDE()`

Function `runDE()` will run differential expression analysis to compare two groups.

By default, `runDE()` will generate pseudobulk matrices using the metadata provided to parameters `replicate_labels` (representing each biological replicate) and, optionally, `split_labels` (representing a further division of the data, often cell types). The pseudobulk matrices will be returned as part of the output of `runDE()`, but they can also be extracted separately using function `getPseudobulk()`.

Alternately, users may opt to run cell-level tests instead of pseudobulk tests (set parameter `pseudobulk`  to "none") or to provide pre-generated pseudobulk matrices (set parameter `pseudobulk`  to "supplied"). See [Advanced Options](<LINK>) for more details.

Function `runDE()` allows the user to run several different methods of differential expression analysis, by adjusting parameters `de_method` and `de_test`:

* `edgeR`: "LRT", "QLF", "exact"
* `DESeq2`: "LRT", "Wald"
* `limma`: "trend", "voom", "wilcox_cpm", "wilcox_cpm_log"
* `presto`: "wilcox_cpm", "wilcox_cpm_log"

By default, `runDE()` will run differential expression analysis using `edgeR` LRT. The precise execution of the differential expression analysis can be further adjusted using parameter `de_params` to pass further parameters to the selected method. This function is parallelized when multiple splits are assessed, so efficiency improves as `n_cores` is increased, up to the number of splits.

```{r, message = TRUE, warning = FALSE}
runDE_output <- runDE(object = sample_data,
                      replicate_labels = "sample",
                      group_labels = "group",
                      split_labels = "cell_type",
                      reference_group = "WT",
                      n_cores = 2)
```

The output is a list containing the following elements:

* `DE_results` Dataframe containing DE results for each feature, by split
* `PB_values` If using pseudobulk data, a list of feature x replicate matri(ces) containing pseudobulk values for each feature, one matrix per split
* `cell_values` Alternately, if using cell-level data, a list of feature x cell matri(ces) containing counts for each feature, one matrix per split
* `metadata` List recording characteristics of the data and runtime
* `parameters` List recording parameter values used

```{r}
head(runDE_output$DE_results)
```

## Step 2: `permuteDE()`

Function `permuteDE()` will take the output from function `runDE()`, permute the group labels and rerun differential expression analysis in the same way across a number of iterations set by parameter `n_iterations`. The parameters are inherited from the `runDE()` function call. For each iteration, the number of significant differentially expressed features is counted, forming a distribution to which the "true" number of significantly differentially expressed features is compared using a permutation test. This permutation test yields a p-value that suggests the probability of observing a number of significant features greater than or equal to the one output by the `runDE()` function if the two groups are not truly different.

The default number of iterations (set by parameter `n_iterations`) is 1000. However, for smaller datasets such as this one where the number of possible combinations of group labels is less than 1000, the permutation test is run using all possible combinations with no repeats. This function is highly parallelized, so efficiency greatly improves as `n_cores` is increased.

```{r, message = TRUE, warning = FALSE}
permuteDE_output <- permuteDE(input = runDE_output,
                              n_cores = 2)
```

The output is a list containing the following elements:

* `permutation_test_results` Dataframe containing the permutation test results by split
* `permutation_DE_summary` Dataframe containing the permutation DE summary metrics by split
* `permutation_DE_results` If parameter `return_all` is `TRUE`, dataframe DE results for each feature, by split, for each iteration (note that this substantially increases the size of the output)
* `metadata` List recording characteristics of the data and runtime
* `parameters` List recording parameter values used

```{r}
head(permuteDE_output$permutation_test_results)
```

```{r}
head(permuteDE_output$permutation_DE_summary)
```

## Plot

permuteDE has several built-in plotting functions to help display the results from functions `runDE()` and `permuteDE()`. All plots are modifiable `ggplot` outputs.

To generate volcano plots for each split assessed by `runDE()` use function `plotVolcano()`. To generate a volcano plot for a specific split or subset of splits, set parameter `use_splits` to the relevant split name(s).

```{r, message = FALSE, warning = FALSE}
# Generate a list of volcano plots
volcano_plots <- plotVolcano(input = runDE_output)
# Display examples
volcano_plots[["Inhibitory Neurons"]] | volcano_plots[["Oligodendrocytes"]]
```

To generate histograms for each permutation test (or a single split), use function `plotHistogram()`. These histograms display the distribution of the number of significant differentially expressed features in each iteration of the permutation test, and a dashed line where the "true" results fall on this distribution.To generate a histogram for a specific split or subset of splits, set parameter `use_splits` to the relevant split name(s).

```{r, message = FALSE, warning = FALSE}
# Generate a list of histograms
histogram_plots <- plotHistogram(input = permuteDE_output)
# Display examples
histogram_plots[["Inhibitory Neurons"]] | histogram_plots[["Oligodendrocytes"]]
```

One way to summarise the results from permuteDE is to color a UMAP by the number of differentially expressed features in each split (in this case, cell type). 

```{r, message = FALSE, warning = FALSE, fig.height = 5}
# Example UMAP showing number of DE features for each split
plotDimReduction(reduction = sample_data@reductions$umap@cell.embeddings,
                 input = permuteDE_output,
                 split_labels = sample_data$cell_type,
                 color_by = "n_sig")
```

In this plot, cell types are grayed out if the permutation test p-value was > 0.05.

```{r, message = FALSE, warning = FALSE, fig.height = 5}
# Example UMAP showing number of DE features for each split that passes permutation testing threshold
plotDimReduction(reduction = sample_data@reductions$umap@cell.embeddings,
                 input = permuteDE_output,
                 split_labels = sample_data$cell_type,
                 color_by = "n_sig",
                 permutation_test_alpha = 0.05)
```

Or we can color by the p-values from the permutation test.

```{r, message = FALSE, warning = FALSE, fig.height = 5}
# Example UMAP showing permutation testing p-value for each split
plotDimReduction(reduction = sample_data@reductions$umap@cell.embeddings,
                 input = permuteDE_output,
                 split_labels = sample_data$cell_type,
                 color_by = "pvalue",
                 palette = "frozen")
```

This function can also be used analogously to `Seurat::FeaturePlot()` to show the values for a single feature across the cells.

```{r, message = FALSE, warning = FALSE, fig.height = 5}
# Example UMAP showing permutation testing p-value for each split
plotDimReduction(reduction = sample_data@reductions$umap@cell.embeddings,
                 feature_values = sample_data@assays$RNA$counts["App",],
                 feature_name = "*App*",
                 color_by = "feature")
```

To plot the results for a single feature, use function `plotFeature()`. By default, this function will apply CPM normalization (using `edgeR::cpm()`), but this can be changed using the parameter `normalization_method`.

```{r, message = FALSE, warning = FALSE}
# Generate a list of plots
feature_plots <- plotFeature(input = runDE_output,
                             feature = "App",
                             use_splits = c("Inhibitory Neurons", "Oligodendrocytes"))
# Display examples
feature_plots[["Inhibitory Neurons"]] | feature_plots[["Oligodendrocytes"]]
```

------

# Advanced Options

------

## Input object types

Function `runDE()` accepts three input object types: Seurat objects, SingleCellExperiment objects, and matrices.

### Seurat

For Seurat objects, the "RNA" assay is used if no input is provided for parameter `use_assay`. If no input is provided for parameter `use_slot`, the "counts" default slot (Seurat v4) or layer (Seurat v5) is used.

Some Seurat v5 objects have multiple layers (e.g., "counts.1", "counts.2", "counts.2"..) with different subsets of cells stored under the same assay. Currently, permuteDE requires a single layer containing all cells in an assay. For these cases, please re-organize the Seurat object prior to running permuteDE.

For Seurat objects, the input to parameters `replicate_labels`, `group_labels`, and `split_labels` may be either a string indicating the name of the metadata column containing the labels or a character vector containing the labels in the same order as the cells.

### SingleCellExperiment

For SingleCellExperiment objects, only the `use_assay` parameter is needed. If not provided, it is set to "counts". Please ensure that the relevant assay matrix includes both row and column names.

For SingleCellExperiment objects, the input to parameters `replicate_labels`, `group_labels`, and `split_labels` may be either a string indicating the name of the metadata column containing the labels or a character vector containing the labels in the same order as the cells.

### Matrix input

For matrix inputs, the provided matrix may be either a cell-level matrix (feature x cell) or a pre-generated pseudobulk matrix (feature x replicate). All permuteDE functions expect raw counts.

If a pre-generated pseudobulk matrix is provided, parameter `pseudobulk` must be set to "supplied".

For matrix inputs, the required input to parameters `replicate_labels`, `group_labels`, and `split_labels` must be provided either:

* As a `data.frame` to parameter `metadata` with column names designated by `replicate_labels`, `group_labels`, and `split_labels`

* Directly to `replicate_labels`, `group_labels`, and `split_labels` as vectors containing the labels in order, corresponding to each column of the matrix

Here's an example of how to apply function `runDE()` to a feature x cell matrix:

```{r, eval = FALSE}
# Extract count matrix from the sample dataset
counts_matrix <- sample_data@assays$RNA$counts
# Extract relevant metadata as a vector
replicates <- sample_data$sample
groups <- sample_data$group
splits <- sample_data$cell_type

# Run 
runDE_output <- runDE(object = counts_matrix,
                      replicate_labels = replicates,
                      group_labels = groups,
                      split_labels = splits,
                      reference_group = "WT",
                      n_cores = 2)
```

Or a pre-generated pseudobulk matrix:

```{r, eval = FALSE}
# Pre-generate pseudobulk matrix
pb_matrix_list <- getPseudobulk(object = sample_data,
                                replicate_labels = "sample",
                                split_labels = "cell_type",
                                n_cores = 2)
pb_matrix <- pb_matrix_list$PB_values[["Oligodendrocytes"]]

# Extract relevant metadata as a vector
replicates <- colnames(pb_matrix)
groups <- sample_data$group[match(replicates, paste0("rep_", sample_data$sample))]

# Run 
runDE_output <- runDE(object = pb_matrix,
                      replicate_labels = replicates,
                      group_labels = groups,
                      reference_group = "WT",
                      pseudobulk = "supplied",
                      n_cores = 1)
```


------

## Complex designs

### Incorporating covariates or multiple factors

To run differential expression analysis accounting for multiple factors/covariates, set parameter `design` in function `runDE()` to a character string providing the desired model formula.

Note that complex model formulas are not compatible with the Wilcoxon rank sum test.

Here is an example comparing the groups within the sample data while additionally correcting for batch effects.

```{r, eval = FALSE}
runDE_output <- runDE(object = sample_data,
                      replicate_labels = "sample",
                      group_labels = "group",
                      split_labels = "cell_type",
                      reference_group = "WT",
                      design = "~ batch + group",
                      n_cores = 2)
```

In most cases, no additional parameter adjustments are needed to run the permutation test with the specified model formula.

```{r, eval = FALSE}
permuteDE_output <- permuteDE(input = runDE_output,
                              n_cores = 2)
```

### Comparisons within the same replicates (e.g., cell type comparisons)

The default usage of `runDE()` does not permit duplicate replicate labels across different comparison groups. Therefore, to compare cell populations wherein the set of replicates is shared across the two groups of interest (e.g., two different cell types or states from the same set of samples), set the parameter `replicate_labels` in function `runDE()` to a concatenation of the replicate labels and the group labels.

```{r, message = TRUE, warning = FALSE}
# Add concatenated metadata column
sample_data$sample_cell_type <- paste(sample_data$sample, sample_data$cell_type, sep = "_")
```

Then, you can run either an unpaired or paired test:

#### Option 1: Unpaired

Here is an example comparing the gene expression between excitatory vs. inhibitory neurons in the sample dataset:

```{r, eval = FALSE}
runDE_output <- runDE(object = sample_data,
                      replicate_labels = "sample_cell_type",
                      group_labels = "cell_type",
                      use_cells = colnames(sample_data)[sample_data$cell_type %in% c("Excitatory Neurons", "Inhibitory Neurons")],
                      reference_group = "Excitatory Neurons",
                      n_cores = 2)

permuteDE_output <- permuteDE(input = runDE_output,
                              n_cores = 2)
```

#### Option 2: Paired

To run a paired comparison, set parameter `design` in function `runDE()` to an additive model formula without an interaction term.

For example, to compare the gene expression between excitatory vs. inhibitory neurons in the sample dataset, our model formula would be `~ sample + cell_type`.

```{r, eval = FALSE}
runDE_output <- runDE(object = sample_data,
                      replicate_labels = "sample_cell_type",
                      group_labels = "cell_type",
                      use_cells = colnames(sample_data)[sample_data$cell_type %in% c("Excitatory Neurons", "Inhibitory Neurons")],
                      reference_group = "Excitatory Neurons",
                      design = "~ sample + cell_type",
                      n_cores = 2)
```

When running function `permuteDE()` on paired comparisons, it is important to provide the correct input to parameter `permute_within`, such that group labels are shuffled separately within each paired set. In this example, `permute_within` should be set to `sample`.

```{r, eval = FALSE}
permuteDE_output <- permuteDE(input = runDE_output,
                              permute_within = "sample",
                              n_cores = 2)
```


------

## permuteDE parameters

### Significance level & multiple comparison correction

The default significance level used by permuteDE is `alpha = 0.05` with multiple comparison correction using FDR = 0.05. The method of multiple comparison correction can be changed using parameter `p_adjust_method` (for permitted values, see `stats::p.adjust.methods`). By default, permuteDE also requires that significant differentially expressed features have a |LFC| > 0.5 (parameter `lfc_threshold`).

### Filters

permuteDE uses various filters to reduce the number of DE comparisons:

*  Parameter `min_cells_per_split` indicates the minimum number of cells within one split. Defaults to 100.
*  Parameter `min_cells_per_replicate` indicates the minimum number of cells within one replicate for each split. Defaults to 10.
*  Parameter `min_replicates_per_split` indicates minimum number of distinct replicates represented within one split. Defaults to 6.
*  Parameter `min_replicates_per_group` indicates the minimum number of distinct replicates represented within each of the two comparison groups. Defaults to 3.
*  Parameter `min_cells_per_feature` indicates the minimum number of cells (within a split) with expression of a feature. Defaults to 10.
*  Parameter `min_prop_cells_per_feature` indicates the minimum proportion of cells (within a split) with expression of a feature. Defaults to 0.1.

### Pseudobulk vs. cell-level tests

We highly encourage users to use pseudobulk differential expression analysis, as it reduces the incidence of false positives. However, permuteDE is also compatible with cell-level tests. To run function `runDE()` without pseudobulking, and treating the cells as the statistical sample size, set parameter `pseudobulk` to "none". This works with any input object type, including cell x feature matrices. Please note that the size of the returned output will be substantially larger, since the cell x feature matrices will be returned as output so that they can be subsequently passed to function `permuteDE()`.

```{r, eval = FALSE}
runDE_output <- runDE(object = sample_data,
                      group_labels = "group",
                      reference_group = "WT",
                      pseudobulk = "none",
                      n_cores = 2)
```

When running cell-level tests, we still recommend that users provide the biological replicate labels to function `permuteDE()` using parameter `permute_by`, such that when the group labels are permuted, cells from each biological replicate are not separated. In other words, the permutation occurs at the biological replicate level, even for cell-level tests. Input to `permute_by` must either be a column name already present in the output under the dataframe `$metadata$group_key` from function `runDE()`, or a vector with values corresponding to each row of that dataframe.

If input is not provided to parameter `permute_by`, the group labels will be permuted across cells regardless of the biological replicate.

```{r, eval = FALSE}
# Grab sample labels for each row of group key
sample_labels <- sample_data@meta.data[runDE_output$metadata$group_key$replicate, "sample"]
# Provide to parameter 'permute_by'
permuteDE_output <- permuteDE(input = runDE_output,
                              permute_by = sample_labels,
                              n_cores = 2)
```


------

## Palettes

Four color palettes are included in this package. They can be called using function `permuteDEpalette()`, and setting parameter `swatch` to `TRUE` will display the palette.

```{r, eval = FALSE}
permuteDEpalette(type = "discrete", 
                  n = 10, 
                  palette = "choir", 
                  swatch = TRUE)
```

```{r, echo = FALSE, fig.height = 1}
palette_choir <- permuteDEpalette(type = "discrete", n = 10, palette = "choir", swatch = TRUE)
palette_archr <- permuteDEpalette(type = "discrete", n = 10, palette = "archr", swatch = TRUE)
palette_inferno <- permuteDEpalette(type = "gradient", n = 10, palette = "inferno", swatch = TRUE)
palette_frozen <- permuteDEpalette(type = "gradient", n = 10, palette = "frozen", swatch = TRUE) 
```

