---
title: "permuteDE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{permuteDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

# Installation

permuteDE is designed to be run on Unix-based operating systems such as macOS and linux.

permuteDE installation currently requires `remotes` and `BiocManager` for installation of GitHub and Bioconductor packages. Run the following commands to install the various dependencies used by CHOIR:

First, install remotes (for installing GitHub packages) if it isn’t already installed:
```{r, eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) install.packages("remotes")
```

Then, install BiocManager (for installing bioconductor packages) if it isn’t already installed:
```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
```

Then, install permuteDE:
```{r, eval = FALSE}
remotes::install_github("corceslab/permuteDE", ref="main", repos = BiocManager::repositories(), upgrade = "never")
```

------

# Introduction

This vignette provides a basic example of how to run permuteDE, a <QQQ>. permuteDE is applicable to <QQQ> and <QQQ> (see [Advanced Options](<LINK>)). Detailed parameter definitions are available under the [Functions](<LINK>) tab.

permuteDE is based on the premise that <QQQ>. The use of permutation testing allows permuteDE to <QQQ>.

permuteDE proceeds in two main steps. First, the user will <QQQ>. Then, <QQQ>.

You’ll need the following packages installed to run this tutorial:
```{r, message = FALSE, warning = FALSE}
library(Seurat)
library(permuteDE)
```

### Example data

To demonstrate how to run permuteDE on a single-cell RNA-seq dataset, we’ll use a small dataset consisting of <QQQ>, originating from an experiment by <QQQ> et al. (<QQQ>), and available through the <Bioconductor package> <QQQ>.

```{r, message = FALSE, warning = FALSE}

```

### Pre-processing

permuteDE takes as input a Seurat, SingleCellExperiment, or matrix. <QQQ> see [Advanced Options](<LINK>)). Note that the exact pre-processing steps used are up to the user.

This tutorial uses a Seurat object; for details related to other object types, see [Advanced Options](<LINK>). <QQQ>

```{r, message = FALSE, warning = FALSE, results = FALSE}

```

------

# Running permuteDE

permuteDE proceeds in two main steps: 

1. Function `runDE()` <QQQ>.
2. Function `permuteDE()` <QQQ>.

## Step-by-step:

### Step 1: `runDE()`

<QQQ>. permuteDE is highly parallelized, so efficiency greatly improves as `n_cores` is increased. <QQQ> using default parameter settings.

```{r, message = TRUE, warning = FALSE, results = "hide"}

```

The output contains <QQQ>. 

### Step 2: `permuteDE()`

<QQQ>. This permutation test yields a p-value that <QQQ>. 

```{r, message = TRUE, warning = FALSE, results = "hide"}

```

The output contains <QQQ>. 

## Plot

<QQQ>. 

```{r}

```

------

# Advanced Options

## Input object types

### Seurat

For Seurat objects, the "RNA" assay is used if no input is provided for parameter `use_assay`. If no input is provided for parameter `use_slot`, the "counts" default slot (Seurat v4) or layer (Seurat v5) is used.

Some Seurat v5 objects have multiple layers (e.g., "counts.1", "counts.2", "counts.2"..) with different subsets of cells stored under the same assay. Currently, permuteDE requires a single layer containing all cells in an assay. For these cases, please re-organize the Seurat object prior to running permuteDE.

### SingleCellExperiment

For SingleCellExperiment objects, only the `use_assay` parameter is needed. If not provided, it is set to "counts". Please ensure that the relevant assay matrix includes both row and column names.

### Matrix input

<QQQ>

## permuteDE parameters

### Significance level & multiple comparison correction

The default significance level used by permuteDE is `alpha = 0.05` with multiple comparison correction (parameter `p_adjust_method`) using FDR = 0.05. By default, permuteDE also requires that significant DEGs have a |LFC| > 0.5 (parameter `lfc_threshold`).

### Filters

permuteDE uses various filters to reduce the number of DE comparisons:

*  Parameter `<QQQ>` indicates the minimum <QQQ>. It defaults to <QQQ>.
